---
title: "ELISA Summary"
output: html_notebook
---

## Read Elisa Data

```{r}
library(readr)
library(tibble)
elisa_data <- read_csv("elisa.csv")
fit_data = head(elisa_data, 7)
fit_data = as_tibble(fit_data)
fit_data
```


## Standard Curve Data

- just performing standard plots here, now doing it as a semi-pro R Dev. Two packages will likely call me that even though there is still much work to be done.

```{r}
#plot(fit_data$conc, fit_data$od)
plot(fit_data$conc, fit_data$od, pch = 21, bg = 'black', ylab = 'od', xlab = 'concentration')
```

## Negative Controls
        
        BLANKS:
        0. Normal vs Chromogen blank.
        1. A completely empty well that has never seen liquid OR a well filled with the ELISA buffer.
        2. Measure minimum background or noise with the normal blank (OD > 0.25 == high noise). Noise from the well or elisa buffer.
        3. Chromogen blank measures noise originating from the substrate or well. How does the original substrate behavior or color contribute to noise? chromogen noise level > normal blank level. 
        
      __SANDWICH__
        
      S0(zero standard or sample/analyte):
      1. Add everything needed for color production except the target analyte or sample (matrix).
      2. Maximum background or noise. S0 noise level > Blank noise level.
      3. Utilized to obtain the null or zero concentration OD value of the std curve.
      
      Negative Matrix Control:
      1. Quantify noise from non-analyte components of the matrix. 
      2. Different from S0 because a matrix (essentially a sample) is added to the mixture.
      
      __COMPETITIVE__
      
      Non-specific binding (NSB):
      1. Quantifies noise from the conjugated enzyme or competitor. 
      2. Maximum noise derived by sample absence vs minimum noise derived by sample presence.
      3. A key point here is NSB is not the expected phenomenom so noise here should not be high (<0.25). This reading should be close to the blank reading OR S0 for sandwich ELISA.
        
        
## Positive Control
      
      B0:
      1. Quantify noise in the absence of a sample. 
      2. Maximum OD obtained relative to standards because of competition absence.
      3. The value is used for adjustments of other values by doing a B/B0 percentage.
      
      
# Quality Control (Incubation Time)

      __COMPETITIVE__
      
      TA (Total activity):
      1. Incubate substrate and conjugated enzyme only followed by a stop solution.
      2. I purport optimal incubation time should be the time it takes to reach the B0 OD value.
      3. 
      





## Intros to Algorithms
